---
title: "Storm Analysis"
author: "Mahesh"
date: "9 September 2016"
output: html_document
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

- [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

#  2. Assignment
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. 

## 2.1 Load Data

```{r cache=TRUE}
storm_data <- read.csv("repdata_data_StormData.csv")
```

### 2.2 Question:

- Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?


```{r}
names(storm_data)

# compute the fatal and injury data by event type and order them 
# (highest to lowest)
fatal <- aggregate(FATALITIES ~ EVTYPE, storm_data, FUN = sum)
fatal <- fatal[order(-fatal$FATALITIES), ]
injury <- aggregate(INJURIES ~ EVTYPE, storm_data, FUN = sum)
injury <- injury[order(-injury$INJURIES), ]

head(injury)
head(fatal)
```

Note that there are 985 Event Types and we will be plotting for top 10 types

```{r}
# bar graphs for top 10 
par(mfrow = c(1, 2), mar = c(14, 4, 4, 4))
barplot(fatal[1:10,]$FATALITIES, las = 3, names.arg = fatal[1:10,"EVTYPE"], 
        main = "Top 10 Events with \nHighest Fatalities", 
        ylab = "Number of fatalities", col = "blue")
barplot(injury[1:10,]$INJURIES, las = 3, names.arg = injury[1:10,"EVTYPE"], 
        main = "Top 10 Events with \nHighest Injuries", 
        ylab = "Number of injuries", col = "blue")

```

### 2.2 Answer:
- 'TORNADO', 'EXCESSIVE HEAT' and 'FLASH FLOOD' EVTYPES produce highest fatalities
- 'TORNADO', 'TSTM WIND' and 'FLOOD' EVTYPES produce highest injuries

### 2.3 Question:
- Across the United States, which types of events have the greatest economic consequences?

```{r}
tail(storm_data$PROPDMG)
tail(storm_data$PROPDMGEXP)
tail(storm_data$CROPDMG)
tail(storm_data$CROPDMGEXP)
```

Notice that the data needs to be computed from the two fields, PROPDMG, PROPDMGEXP 
and from CROPDMG , CROPDMGEXP

```{r}
#from data find the multipliers for k,m,B and other values 
damage_multiplier <- data.frame(
                        PROPDMGEXP = c ("-","?","+","","0","1","2","3","4",
                                        "5","6","7","8","h","H",
                                        "K","m","M","B"),
                        MLTPLYR = c(0,0,0,0,1,10,100,1000,10000,
                                    100000,1000000,10000000,100000000,
                                    100,100,1000,1000000,1000000,1000000000))
damage_multiplier2 <- data.frame(
                CROPDMGEXP = c ("?","","0","2","k","K","m","M","B"),
                MLTPLYR2 = c(0,0,1,100,1000,1000,1000000,1000000,1000000000))

# merge the data to get the MLTYPLR, MLTYPLYR2 added to the data set

storm_data <- merge(storm_data, damage_multiplier, by = "PROPDMGEXP")

# compute the property value by 
storm_data$PROPDMGVAL <- storm_data$PROPDMG * storm_data$MLTPLYR

storm_data <- merge(storm_data, damage_multiplier2, by = "CROPDMGEXP")

storm_data$CROPDMGVAL <- storm_data$CROPDMG * storm_data$MLTPLYR2
```

Aggregate the values from PROPDMGVAL and CROPDMGVAL by EVTYPE

```{r}
# compute the property and crop damage data by event type and order them 
# (highest to lowest)
propdmg <- aggregate(PROPDMGVAL ~ EVTYPE, storm_data, FUN = sum)
propdmg <- propdmg[order(-propdmg$PROPDMGVAL), ]

cropdmg <- aggregate(CROPDMGVAL ~ EVTYPE, storm_data, FUN = sum)
cropdmg <- cropdmg[order(-cropdmg$CROPDMGVAL), ]
```

Note that there are 985 Event Types and we will be plotting for top 10 types

```{r}
# bar graphs for top 10 
par(mfrow = c(1, 2), mar = c(14, 4, 4, 4))

barplot(propdmg[1:10,]$PROPDMGVAL/(10^9), las = 3, 
        names.arg = propdmg[1:10,"EVTYPE"], 
        main = "Events with Highest \nProperty damages", 
        ylab = "Value of Damage in Billions", col = "blue")
barplot(cropdmg[1:10,]$CROPDMGVAL/(10^9), las = 3, 
        names.arg = cropdmg[1:10,"EVTYPE"], 
        main = "Events with Highest \nCrop damages", 
        ylab = "Value of Damage in Billions", col = "blue")
```

### 2.3 Answer:
- 'FLOOD', 'HURRICANE/TYPHOON' and 'TORNADO' EVTYPES produce highest property damages
- 'DROUGHT', 'FLOOD' and 'RIVER FLOOD' EVTYPES produce highest crop damages
